---
title: "Migrating from RClickhouse 0.6.x to 2.0"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Migrating from RClickhouse 0.6.x to 2.0}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

RClickhouse 2.0 is a ground-up rewrite. Instead of bundling the C++ ClickHouse
client and marshalling every value through a custom Rcpp layer, it now speaks to
ClickHouse entirely through **Apache Arrow** — either the native **Arrow Flight
SQL** protocol (via ADBC) or the **HTTP** interface with the `ArrowStream`
format. The package contains no compiled code anymore.

This vignette summarises what changed for existing users.

## The short version

Most code keeps working. The main change is **how you connect**: there is no
more native binary protocol on port `9000`. Pick a transport instead.

```{r}
library(DBI)

# Old (0.6.x): native protocol, port 9000
con <- dbConnect(RClickhouse::clickhouse(), host = "db", port = 9000)

# New (2.0): Arrow Flight SQL (default), port 9090
con <- dbConnect(RClickhouse::clickhouse(), host = "db")

# New (2.0): HTTP interface, port 8123
con <- dbConnect(RClickhouse::clickhouse(), host = "db", transport = "http")
```

Everything downstream — `dbWriteTable()`, `dbGetQuery()`, `dbReadTable()`,
`tbl()` + dplyr verbs — is unchanged.

## Choosing a transport

| | `"flightsql"` (default) | `"http"` |
|---|---|---|
| Protocol | Arrow Flight SQL (ADBC) | HTTP + `ArrowStream` |
| Default port | 9090 | 8123 (8443 with TLS) |
| Extra install | `adbcflightsql` (R-multiverse) | none (pure R) |
| Server requirement | `arrowflight_port` enabled (recent ClickHouse; not on ClickHouse Cloud) | works everywhere |

To use Flight SQL, install the driver once (it is not on CRAN):

```{r}
install.packages("adbcflightsql", repos = "https://community.r-multiverse.org")
```

If your server does not expose the Arrow Flight interface (for example
ClickHouse Cloud), use `transport = "http"`.

## Deprecated connection arguments

These still work but emit a deprecation warning and are otherwise ignored or
remapped:

| Old argument | Status |
|---|---|
| `db = ` | renamed to `dbname = ` |
| `compression = ` | ignored (Arrow negotiates compression) |
| `Int64 = ` | ignored (Int64/UInt64 always map to `bit64::integer64`) |
| `toUTF8 = ` | ignored (strings are always UTF-8) |
| `port = 9000` | warns; use the Flight SQL (9090) or HTTP (8123) port |

## Types

Type conversion is now done by Arrow/nanoarrow rather than the old hand-written
mapping. In particular:

* `Int64` / `UInt64` columns come back as `bit64::integer64`.
* Strings are UTF-8.
* `Date` / `DateTime` map to `Date` / `POSIXct`.

`dbColumnInfo()` now reports the R type of each result column rather than the
ClickHouse type string used internally by the old C++ layer.

## dbplyr

The dbplyr backend is unchanged in behaviour: ClickHouse-specific translations
(`^` to `pow()`, `var()`/`sd()` to `varSamp()`/`stddevSamp()`, case-sensitive
function names via `dbplyr_case_sensitive()`) are retained.
```
